### 2013-10-21 article.tex
Expanded section on WARP. Added footnote references.

\usepackage{color}
\usepackage{float}
\usepackage{mdwlist}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{verbatim}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{listings}

@@

% INTRODUCTION
\section*{Introduction} % The \section*{} command stops section numbering
\addcontentsline{toc}{section}{\hspace*{-\tocsep}Introduction} % Adds this section to the table of contents with negative horizontal space equal to the indent for the numbered sections
\label{sec:introduction}
% Background and Challenge
%	Vilket sammanhang utgör studiens bakgrund?
%	Vilka utmaningar omfattar studien?
When developing \textit{GPU}-kernels relatively small changes may induce large deviations in performance, due to massively parallelized instruction sets and architectural differences in-
between on-chip hardware. Therefore, it may be desirable for the developer to be able to view the data he or she is modifying on the graphics card - a possibility often limited in terms of \textit{GPGPU}-
technologies.\\
% Approach and Result
%	Vilken ansats tillämpas i studien?
The increased utilization of \textit{GPGPU} has brought forth the need of more extensive debugging-possibilities involving access of data that may be hard to retrieve from the hardware - possibly due to architectural differences in-between chip manufacturers. A preferred solution to this problem has been to emulate such \textit{GPU}-kernels on the \textit{CPU} - in exchange for substantial performance-losses. Other resons to emulate \textit{GPU}-kernels may concern pre-silicon development - that is, development for hardware not yet existant, when hardware is busy, or otherwize unavailable. This study comprises an investigation into the performance of software emulation of hardware accelerated \textit{GPGPU}-kernels, by the means of analyzing several software rasterizers.\\
\\
Furthermore, this study concerns inquiry into the \textit{DirectCompute}-framework on the \textit{Windows}-platform, analyzing the performance of a \textit{GPGPU}-kernel on-chip, using the \textit{DirectX} standard software rasterizer, and utilizing the \textit{DirectX~11.1}-addition Microsoft~\textit{WARP}\footnote{\textit{Windows~Advanced~Rasterization~Platform}.}-technology - which promises high-speed software emulation in exchange for some computational precision\footnote{See Microsoft's \textit{WARP~Guide}: \url{http://bit.ly/19hiwrZ}}. \\
%	Vilka resultat gav studien?
%\\
%This study concludes that the performance loss of emulated \textit{GPGPU}-kernel may be considered negligible in exchange for possible extended possibilities of data extraction and increased granularity. As such, Microsoft~\textit{WARP}-technology may be considered feasible for use in industry rasterization\footnote{Such a use-case might be if hardware is unavailable, not suffiecient or busy - as suggested by Microsoft. Link to Microsofts page on WARP here!}.\\
\\
% Outline and Conclusion
As such, the study concerns the fields of simulation, emulation and \textit{GPU}-technologies - respectively, with the purpose of facilitating debugging and profiling of \textit{GPU}-kernels, whilst maintaining acceptable performance.
%	Vilket upplägg har resterande delar av matrialet?

@@

\subsection{Subjects~of~Study}
\label{sec:contribution:subjectofstudy}
% Describe these modes and what they entail:
The \textit{GPU}-kernels described in Section~\ref{sec:contribution:method} are comprised of \textit{HLSL}-syntax and are compiled \& executed using Microsoft~\textit{DirectCompute}. These kernels are run using three types of hardware~acceleration with varying strategies of \textit{GPU}-kernel emulation. These are comprised of the following:
\paragraph{Hardware-Acceleration~(GPU)}
	The execution of a \textit{DirectCompute}-kernel on a graphics card. Thus, this case will act as a reference for the emulated subjects.\\
	Expected high performance.
\paragraph{Software~Rasterization~(CPU)}
	The emulated execution of a \textit{DirectCompute}-kernel using the \textit{DirectX}~Reference~Device~Driver.\\
	Expected poor performance.
\paragraph{Windows~Advanced~Rasterization~Platform~(CPU)}
	The emulated execution of a \textit{DirectCompute}-kernel using a special software rasterizer devised by Microsoft in their latest revision of the \textit{DirectX}-framework.\\	
	Expectations unclear, but expected to perform better than standard software rasterization.\footnote{Insert reference to microsoft warp-msdn-page and explain what WARP promises and stuff about optimization for gpus.}\\

@@

\subsection{Kernels}
\label{sec:contribution:kernels}
In addition to the subject drivers mentioned in Section~\ref{sec:contribution:subjectofstudy}, two kernels with varying lavel of optimization have been examined\footnote{Why?}. These kernels are presented below.
\paragraph{Matrix~mult.~w.~Blocks}
	A kernel producing $AB=C$ from two given matrices, writing back $C$ for further analysis.\\
	The kernel is executed with one thread for each element in the square matrices, and likewize produce a lone element of the resulting matrix thread-wize.\\
	Execution is performed in blocks of $16x16$ threads, as this number was found to be the block dimension performing optimally whilst hardware~accelerated on the \textit{GPU}\footnote{Make sure of this!}. 
\paragraph{Matrix~mult.~w.~Blocks~\&~Shared~Memory}
	Likewize as with the previous kernel, but further optimized to utilize \textit{shared~memory} in order to reduce time-consuming reading of \textit{global~memory}\cite[p.~77-93]{Kirk:2010:PMP:1841511}.\\
	This kernel is presented as a test-case due to the preconditions of \textit{WARP} - stating that a kernel optimized for \textit{GPU}-execution is likewize optimized for execution with \textit{WARP}. Hence, we investigate a more optimized kernel to see whether or not this behaviour may be replicated in the experiment.\\

@@

\paragraph{analytics}
\textit{analytics} is a smaller utility developed to compose data surrounding possible precisional deviations in-between matrices $C$ and $Ref$. \textit{analytics} compiles the minimum- and maximum-deviation encountered, as well as to calculate the standard deviation of said precisional deviation. In turn, \textit{analytics} outputs this information to an intermediate file.\\
\textit{analytics} is written in \texttt{C++}.\\
\\
\noindent
These three applications are, in turn, run as subprocesses in a script specifying the various configurations and number of times to run each program. This script, written in \textit{Python}, then compiles the assorted results of these applications and outputs a range of files suitably formatted for interpretation by \textit{Gnuplot}.\\
More information surrounding how to run these programs can be found via the source repository (see Acknowledgments).

@@

\subsubsection{Equipment}
\label{sec:contribution:equipment}
The results presented in this study is gathered from experiments performed on a system with the following specifications:
\begin{description*}
	\item[CPU]	Intel Q9550 Quad Core 2.83GHz
	\item[GPU]	ATI Radeon HD 5800
	\item[OS]	Windows 8.0
\end{description*}

@@

\begin{table}[h]
\begin{center}
\begin{tabular}{r|r|r|r|}
	\cline{2-3}
							& \multicolumn{1}{|c|}{\textbf{BASIC}} & \multicolumn{1}{|c|}{\textbf{TILE}}	\\ \hline
	\multicolumn{1}{|l|}{\textbf{HARD}}	& $1.16$			& $0.24$ 	& $-79.3\%$    				\\ \hline
	\multicolumn{1}{|l|}{\textbf{SOFT}}	& $11610.02$		& $9866.40$	& $-15.0\%$     				\\ \hline
	\multicolumn{1}{|l|}{\textbf{WARP}}	& $15.31$			& $18.97$	& $+23.9\%$     				\\ \hline
\end{tabular}
\label{tab:contribution:results:summaryint}
\caption{Average exectution time (ms) of a $200x200$ integer matrix.}
\end{center}
\end{table}

\begin{table}[h]
\begin{center}
\begin{tabular}{r|r|r|r|}
	\cline{2-3}
							& \multicolumn{1}{|c|}{\textbf{BASIC}} & \multicolumn{1}{|c|}{\textbf{TILE}}		\\ \hline
	\multicolumn{1}{|l|}{\textbf{HARD}}	& $0.77$			& $0.22$		& $-71.4\%$    				\\ \hline
	\multicolumn{1}{|l|}{\textbf{SOFT}}	& $10247.03$		& $10909.88$	& $+6.5\%$     				\\ \hline
	\multicolumn{1}{|l|}{\textbf{WARP}}	& $14.08$			& $17.44$		& $+23.9\%$    				\\ \hline
\end{tabular}
\label{tab:contribution:results:summaryfloat}
\caption{Average exectution time (ms) of a $200x200$ floating point matrix.}
\end{center}
\end{table}

@@

\subsubsection{Computational~Precision}
\label{sec:contribution:results:computationalprecision}
\begin{table}
\begin{center}
	\begin{tabular}{|l|l|l|}
		\hline
		Minimum & Maximum & Standard \\ \hline
		0.0     & ~0.01   & ~0.0025            \\ \hline
	\end{tabular}
\label{tab:contribution:results:computationalprecision:avgprecision}
\caption {Average precisional deviations in floating point operations.}
\end{center}
\end{table}

Integer had no loss of precision whatsoever.
All floating point experiments experienced a minimum precisional deviation of 0, a maximum precisional deviation of ~0.01 and a standard precisional deviation of ~0.0025. Experiments indicated no increased performance loss in-between the different driver types.

@@

% ACKNOWLEDGMENTS
\section*{Acknowledgments} % The \section*{} command stops section numbering
\addcontentsline{toc}{section}{\hspace*{-\tocsep}Acknowledgments} % Adds this section to the table of contents with negative horizontal space equal to the indent for the numbered sections
\label{sec:acknowledgments}
I wish to express my gratitude to my fellow students, colleagues and close friends Bob, Benny and no-one for their work, co-operation and positive attitude throughout the execution of this experiment. \\
\\
The source code used during this study is freely available on \textit{GitHub}\footnote{\textit{GitHub}~Repository:~\url{http://bit.ly/1c4wd20}.}, along with a guide on how to compile and run the solution in order to replicate the experiment.

>>>>

\usepackage{color}
\usepackage{float}
\usepackage{mdwlist}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{verbatim}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{footmisc}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{listings}

@@

% INTRODUCTION
\section*{Introduction} % The \section*{} command stops section numbering
\addcontentsline{toc}{section}{\hspace*{-\tocsep}Introduction} % Adds this section to the table of contents with negative horizontal space equal to the indent for the numbered sections
\label{sec:introduction}
% Background and Challenge
%	Vilket sammanhang utgör studiens bakgrund?
%	Vilka utmaningar omfattar studien?
When developing \textit{GPU}-kernels relatively small changes may induce large deviations in performance, due to massively parallelized instruction sets and architectural differences in-
between on-chip hardware. Therefore, it may be desirable for the developer to be able to view the data he or she is modifying on the graphics card - a possibility often limited in terms of \textit{GPGPU}-
technologies.\\
% Approach and Result
%	Vilken ansats tillämpas i studien?
The increased utilization of \textit{GPGPU} has brought forth the need of more extensive debugging-possibilities involving access of data that may be hard to retrieve from the hardware - possibly due to architectural differences in-between chip manufacturers. A preferred solution to this problem has been to emulate such \textit{GPU}-kernels on the \textit{CPU} - in exchange for substantial performance-losses. Other resons to emulate \textit{GPU}-kernels may concern pre-silicon development - that is, development for hardware not yet existant, when hardware is busy, or otherwize unavailable. This study comprises an investigation into the performance of software emulation of hardware accelerated \textit{GPGPU}-kernels, by the means of analyzing several software rasterizers.\\
\\
Furthermore, this study concerns inquiry into the \textit{DirectCompute}-framework on the \textit{Windows}-platform, analyzing the performance of a \textit{GPGPU}-kernel on-chip, using the \textit{DirectX} standard software rasterizer, and utilizing the \textit{DirectX~11.1}-addition Microsoft~\textit{WARP}\footnote{\textit{Windows~Advanced~Rasterization~Platform}.}-technology - which promises high-speed software emulation in exchange for some computational precision\footnote{\label{lab:warpguide}Microsoft's \textit{WARP~Guide}: \url{http://bit.ly/19hiwrZ}. Retrieved: 21-10-2013.}. \\
%	Vilka resultat gav studien?
%\\
%This study concludes that the performance loss of emulated \textit{GPGPU}-kernel may be considered negligible in exchange for possible extended possibilities of data extraction and increased granularity. As such, Microsoft~\textit{WARP}-technology may be considered feasible for use in industry rasterization\footnote{Such a use-case might be if hardware is unavailable, not suffiecient or busy - as suggested by Microsoft. Link to Microsofts page on WARP here!}.\\
\\
% Outline and Conclusion
As such, the study concerns the fields of simulation, emulation and \textit{GPU}-technologies - respectively, with the purpose of facilitating debugging and profiling of \textit{GPU}-kernels, whilst maintaining acceptable performance.
%	Vilket upplägg har resterande delar av matrialet?

@@

\subsection{Subjects~of~Study}
\label{sec:contribution:subjectofstudy}
% Describe these modes and what they entail:
The \textit{GPU}-kernels described in Section~\ref{sec:contribution:method} are comprised of \textit{HLSL}-syntax and are compiled \& executed using Microsoft~\textit{DirectCompute}. These kernels are run using three types of acceleration~technologies with varying strategies of \textit{GPU}-kernel emulation. These are comprised of the following:
\paragraph{Hardware-Acceleration~(GPU)}
	The execution of a \textit{DirectCompute}-kernel on a graphics card. This is the common case, and involves no emulation of the kernel. Thus, this case will act as a reference for the emulated subjects.\\
	During hardware~acceleration on the \textit{GPU}, one may expect high performance.
\paragraph{Software~Rasterization~(CPU)}
	The emulated execution of a \textit{DirectCompute}-kernel using the \textit{DirectX}~Reference~Device~Driver.\\
The Device~Driver was developed for the purpose of testing and debugging, and is - although it does support some \textit{CPU}-optimizations - not intended to be used in retail applications\footnote{\label{lab:drivertypes}Microsoft's reference on \textit{DirectX}~Driver~Types: \url{http://bit.ly/1b63NiX}. Retrieved: 21-10-2013.}.\\
	As the Reference~Driver is designed for the purpose of accuracy, rather than speed, one may expect poor performance.
\paragraph{Windows~Advanced~Rasterization~Platform~(CPU)}
	The emulated execution of a \textit{DirectCompute}-kernel using a special software rasterizer devised by Microsoft in their latest revision of the \textit{DirectX}-framework.\\
	The driver is based off the \textit{DirectX}~Reference~Driver and uses thread pooling to distribute tasks effectively on the \textit{CPU}, along with grouping execution in batches for optimum performance. Microsoft describes \textit{WARP} as a high-performance software~rasterizer, and recommends using the driver for retail applications, such as 'casual~games' or 'advanced rendering games'. For more information surrounding the \textit{WARP}-driver, see Microsoft's \textit{WARP~Guide}\footref{lab:warpguide}.\\
As there are, as of yet, few studies performed on Microsoft~\textit{WARP} expectations are unclear, but expected to perform better than standard software rasterization.\\
% May one encounter computational precision issues with WARP?

@@

\subsection{Kernels}
\label{sec:contribution:kernels}
In addition to the subject drivers mentioned in Section~\ref{sec:contribution:subjectofstudy}, two kernels with varying lavel of optimization have been examined\footnote{Why?}. These kernels are presented below.
\paragraph{Matrix~mult.~w.~Thread~Blocks}
	A kernel producing $AB=C$ from two given matrices, writing back $C$ for further analysis.\\
	The kernel is executed with one thread for each element in the square matrices, and likewize produce a lone element of the resulting matrix thread-wize.\\
	Execution is performed in blocks of $16x16$ threads, as this number was found to be the block dimension performing optimally whilst hardware~accelerated on the \textit{GPU}\footnote{Make sure of this!}. 
\paragraph{Matrix~mult.~w.~Thread~Blocks~\&~Shared~Memory}
	Likewize as with the previous kernel, but further optimized to utilize \textit{shared~memory} in order to reduce time-consuming reading of \textit{global~memory}\cite[p.~77-93]{Kirk:2010:PMP:1841511}.\\
	This kernel is presented as a test-case due to the preconditions of \textit{WARP} - stating that a kernel optimized for \textit{GPU}-execution is likewize optimized for execution with \textit{WARP}\footref{lab:warpguide}. Hence, we investigate a more optimized kernel to see whether or not this behaviour may be replicated in the experiment.\\

@@

\paragraph{analytics}
\textit{analytics} is a smaller utility developed to compose data surrounding possible precisional deviations in-between matrices $C$ and $Ref$. \textit{analytics} compiles the minimum- and maximum-deviation encountered, as well as to calculate the standard deviation of said precisional deviation. In turn, \textit{analytics} outputs this information to an intermediate file.\\
\textit{analytics} is written in \texttt{C++}.\\
\\
\noindent
These three applications are, in turn, run as subprocesses in a script specifying the various configurations and number of times to run each program. This script, written in \textit{Python}, then compiles the assorted results of these applications and outputs a range of files suitably formatted for interpretation by \textit{Gnuplot}.\\
\\
The source code manufactured for the sake of this study is freely available via \textit{GitHub}\footnote{\label{lab:github}\textit{GitHub}~Repository:~\url{http://bit.ly/1c4wd20}. Retrieved: 21-10-2013.}, along with a guide on how to compile and run the solution in order to replicate the experiment. Furthermore, the results collected and used throughout this study is also available for download, and may be acquired for further analysis.

@@

\subsubsection{Equipment}
\label{sec:contribution:equipment}
The results presented in this study is gathered from experiments performed on a system with the following specifications:
\begin{description*}
	\item[CPU]	Intel Q9550 Quad Core 2.83GHz
	\item[GPU]	ATI Radeon HD 5800
	\item[OS]	Windows 8.0
\end{description*}
The study has been performed on this system as Microsoft claims that the \textit{WARP}~driver performs best on modern quad-core \textit{CPU}s\footref{lab:warpguide}.

@@

\begin{table}[h]
\begin{center}
\begin{tabular}{r|r|r|r|}
	\cline{2-3}
							& \multicolumn{1}{|c|}{\textbf{BASIC}} & \multicolumn{1}{|c|}{\textbf{TILE}}	\\ \hline
	\multicolumn{1}{|l|}{\textbf{HARD}}	& $1.16$			& $0.24$ 	& $-79.3\%$    				\\ \hline
	\multicolumn{1}{|l|}{\textbf{SOFT}}	& $11610.02$		& $9866.40$	& $-15.0\%$     				\\ \hline
	\multicolumn{1}{|l|}{\textbf{WARP}}	& $15.31$			& $18.97$	& $+23.9\%$     				\\ \hline
\end{tabular}
\label{tab:contribution:results:summaryint}
\caption{Average exectution time in milliseconds of a $200x200$ integer matrix.}
\end{center}
\end{table}

\begin{table}[h]
\begin{center}
\begin{tabular}{r|r|r|r|}
	\cline{2-3}
							& \multicolumn{1}{|c|}{\textbf{BASIC}} & \multicolumn{1}{|c|}{\textbf{TILE}}		\\ \hline
	\multicolumn{1}{|l|}{\textbf{HARD}}	& $0.77$			& $0.22$		& $-71.4\%$    				\\ \hline
	\multicolumn{1}{|l|}{\textbf{SOFT}}	& $10247.03$		& $10909.88$	& $+6.5\%$     				\\ \hline
	\multicolumn{1}{|l|}{\textbf{WARP}}	& $14.08$			& $17.44$		& $+23.9\%$    				\\ \hline
\end{tabular}
\label{tab:contribution:results:summaryfloat}
\caption{Average exectution time in milliseconds of a $200x200$ floating point matrix.}
\end{center}
\end{table}

@@

\subsubsection{Computational~Precision}
\label{sec:contribution:results:computationalprecision}
\begin{table}
\begin{center}
	\begin{tabular}{|r|r|r|}
		\hline
		\textbf{Minimum} 	& \textbf{Maximum} 	& \textbf{Standard} 	\\ \hline
		0.0     			& ~0.01   			& ~0.0025            		\\ \hline
	\end{tabular}
\label{tab:contribution:results:computationalprecision:avgprecision}
\caption {Average precisional deviations in floating point operations.}
\end{center}
\end{table}

%http://msdn.microsoft.com/en-us/library/windows/desktop/cc308050%28v=vs.85%29.aspx
Integer had no loss of precision whatsoever.
All floating point experiments experienced a minimum precisional deviation of 0, a maximum precisional deviation of ~0.01 and a standard precisional deviation of ~0.0025. Experiments indicated no increased performance loss in-between the different driver types.

@@

% ACKNOWLEDGMENTS
\section*{Acknowledgments} % The \section*{} command stops section numbering
\addcontentsline{toc}{section}{\hspace*{-\tocsep}Acknowledgments} % Adds this section to the table of contents with negative horizontal space equal to the indent for the numbered sections
\label{sec:acknowledgments}
I wish to express my gratitude to my fellow students, colleagues and close friends Bob, Benny and no-one for their work, co-operation and positive attitude throughout the execution of this experiment.
